---
title: "Predicting pedigree breeding values"
subtitle: "Practical 4: BLUP, shrinkage and prediction uncertainty"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
---

## Learning objectives

Distinguish breeding values from phenotypes and fixed group effects; calculate BLUP under a pedigree model; explain shrinkage and information from relatives; and distinguish prediction reliability from observed prediction accuracy.

Continue from [Practical 3](mouse_genetic_parameters.qmd). You need base R, `qgg` and the [shared helper script](../scripts/mouse_practical_helpers.R). This practical uses the same mouse data and founder base. Data checks and preparation are repeated so the page runs independently.

## 1. Fit the pedigree model

```{r}
#| label: prepare-blup
source("../scripts/mouse_practical_helpers.R")
data <- mouse_pedigree_data()
mouse <- data$mouse
ids <- data$ids
A <- mouse_relationship(data$pedigree)
K <- A[ids, ids, drop=FALSE]
X <- model.matrix(~sex+reps, data=mouse)
y <- mouse$BW
names(y) <- ids
fitA <- mouse_fit(y, X, K, "A")
fitA$theta
fitA$niter
VA <- unname(fitA$theta["A"])
VE <- unname(fitA$theta["E"])
```

The model is $y=Xb+u+e$, with $\operatorname{Var}(u)=A\sigma_A^2$ and independent residual variance $I\sigma_E^2$. Estimated variance components give **empirical BLUP**: prediction conditional on the fitted components. Shared-litter/maternal effects are absent from this model, so model adequacy matters.

## 2. Separate the parts of a phenotype prediction

The fitted prediction is $\hat y_i=x_i^T\hat b+\hat u_i$. The fixed part describes sex and replicate contrasts; $\hat u_i$ is the predicted additive genetic deviation relative to the pedigree base. Residuals are $y_i-\hat y_i$.

```{r}
#| label: prediction-table
u_hat <- drop(fitA$g[,"A"])
predictions <- data.frame(id=ids, phenotype=y,
                          fixed_part=drop(X %*% fitA$b),
                          breeding_value=u_hat,
                          predicted_phenotype=drop(fitA$predicted),
                          residual=drop(fitA$e), row.names=NULL)
stopifnot(max(abs(predictions$predicted_phenotype-
                  predictions$fixed_part-u_hat)) < 1e-8,
          max(abs(y-predictions$predicted_phenotype-predictions$residual)) < 1e-8)
knitr::kable(head(predictions), digits=3)
```

**Question 1.** Why can two mice with similar body weights receive different breeding-value predictions?

::: {.callout-note collapse="true" title="Suggested answer"}
Their sex/replicate adjustments and relatives' records may differ. Phenotype ranking mixes group means, additive effects and residual sources. BLUP combines adjusted records using the relationship and variance model. A higher prediction concerns the defined trait and base; it does not automatically imply a better multivariate breeding objective.
:::

## 3. Calculate BLUP directly

For $V=A\sigma_A^2+I\sigma_E^2$,

$$\hat b=(X^TV^{-1}X)^{-1}X^TV^{-1}y,\qquad
\hat u=\sigma_A^2 A V^{-1}(y-X\hat b).$$

The following base-R calculation verifies the package prediction at its fitted variance components.

```{r}
#| label: direct-blup
V <- VA*K + diag(VE, length(y))
Vi <- chol2inv(chol(V))
ViX <- Vi %*% X
Cb <- solve(crossprod(X, ViX))
b_direct <- Cb %*% crossprod(X, Vi %*% y)
u_direct <- drop(VA*K %*% Vi %*% (y-X %*% b_direct))
stopifnot(max(abs(u_direct-u_hat)) < 1e-7,
          max(abs(b_direct-fitA$b)) < 1e-7)
```

**Question 2.** What happens when residual variance is large relative to additive variance?

::: {.callout-note collapse="true" title="Suggested answer"}
A record carries less information about its underlying breeding value. Predictions generally shrink more toward the model's reference expectation. The amount depends on the whole information structure, not only one phenotype.
:::

### An own-record illustration

For one noninbred, unrelated individual with known fixed mean, $u\sim N(0,\sigma_A^2)$ and $e\sim N(0,\sigma_E^2)$. Its prediction from an adjusted phenotype $y^*$ is $\hat u=h^2y^*$, with $h^2=\sigma_A^2/(\sigma_A^2+\sigma_E^2)$. This scalar illustration does not replace the mouse pedigree analysis.

```{r}
#| label: fig-shrinkage
#| fig-cap: "Own-record shrinkage at different assumed heritabilities; the group mean is known in this illustration."
adjusted_record <- seq(-10,10,length.out=101)
plot(adjusted_record, 0.2*adjusted_record, type="l", ylim=c(-10,10),
     xlab="Adjusted phenotype", ylab="Predicted breeding value",
     col="#247BA0", lwd=2)
lines(adjusted_record, 0.8*adjusted_record, col="#B65C39", lwd=2)
abline(0,1,lty=2)
legend("topleft", c("h²=0.2", "h²=0.8", "Unshrunk record"),
       col=c("#247BA0","#B65C39","black"), lty=c(1,1,2), bty="n")
stopifnot(0.2*5==1, 0.8*5==4)
```

## 4. Explore pedigree predictions

```{r}
#| label: fig-pedigree-predictions
#| fig-cap: "Predicted breeding values and model residuals for recorded mice. Association with input phenotypes is not validation accuracy."
#| fig-width: 9
#| fig-height: 5
local({
  old <- par(mfrow=c(1,2)); on.exit(par(old))
  plot(y, u_hat, pch=16, cex=0.5, col="#247BA0",
       xlab="Observed body weight", ylab="Pedigree breeding value")
  hist(drop(fitA$e), breaks=25, col="#86B7A3", border="white",
       xlab="Model residual", main="Residual distribution")
})
```

A residual histogram can identify patterns worth investigating; it cannot prove independence or eliminate omitted family effects. Correlation with a phenotype already used for fitting is an in-sample association, not evidence of predictive accuracy.

## 5. Quantify model-based uncertainty

With unknown fixed effects accounted for, define

$$P=V^{-1}-V^{-1}X(X^TV^{-1}X)^{-1}X^TV^{-1}.$$

The prediction-error covariance at supplied variance components is

$$C_u=\sigma_A^2 A-\sigma_A^4 APA.$$

Its diagonal contains prediction-error variances (PEV). A model-based reliability is $1-PEV_i/(A_{ii}\sigma_A^2)$; its square root is often called model-based accuracy. These plug-in quantities ignore uncertainty in the estimated variance components and do not provide empirical validation against true breeding values.

```{r}
#| label: prediction-reliability
P <- Vi - ViX %*% Cb %*% t(ViX)
# Compute the diagonal without storing another full covariance matrix.
pev <- VA*diag(K) - VA^2*rowSums((K %*% P)*K)
reliability <- 1-pev/(VA*diag(K))
stopifnot(all(pev >= -1e-8), all(reliability >= -1e-8),
          all(reliability <= 1+1e-8))
summary(reliability)
predictions$reliability <- reliability
knitr::kable(head(predictions[order(-predictions$breeding_value),
                               c("id","breeding_value","reliability")]),digits=3)
# Independent scalar check: own record, known mean, VA=2, VE=3.
scalar_pev <- 2-2^2/(2+3)
stopifnot(abs(1-scalar_pev/2 - 2/(2+3)) < 1e-12)
```

**Question 3.** Does the highest breeding-value prediction necessarily have the greatest reliability?

::: {.callout-note collapse="true" title="Suggested answer"}
No. Predicted merit and precision answer different questions. Reliability depends on information, relationships and variance assumptions. True breeding values are unavailable for these mice; their empirical prediction accuracy cannot be read from this table.
:::

## Exercises

1. Compare ranking by phenotype and by predicted breeding value. How do group adjustments affect interpretation?
2. Identify a mouse with relatively low reliability. Describe what additional information might help.
3. Explain why retaining only the largest predicted values does not evaluate response to selection.

::: {.callout-note collapse="true" title="Suggested solutions"}
```{r}
head(predictions[order(-predictions$phenotype),c("id","phenotype","breeding_value")])
head(predictions[order(predictions$reliability),c("id","breeding_value","reliability")])
```
More informative own/relative records can improve precision under an appropriate model. Selection response also depends on mating, inheritance, the objective and the next generation's environment. `gsel` owns decision policies and `gsim` biological simulation; neither fits this pedigree evaluation.
:::

## Next practical

[Practical 5](mouse_genomic_breeding_values.qmd) replaces pedigree expectations with a genomic relationship model and separates agreement between predictions from validation on held-out records.
